{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-07-28 thesis/landslide.tex
ok i think retry set impl is final, so write about it

{\bf Further pruning.}
In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
For example, consider the execution sequence
in Figure~\ref{fig:sleepsets}(a).
After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
Figure~\ref{fig:sleepsets}(b).
However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
so from this branch,
it next tags $I_{c'1}$, causing
Figure~\ref{fig:sleepsets}(c).
to be tested next, which is regrettably equivalent to the original interleaving
($t_a$ and $t_b$ having been reordered all told, which are independent).
To recover our forgotten original purpose,
which was already fulfilled by testing the second interleaving,
DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
for an already-explored interleaving beginning with the target thread.
If such exists,
%then because of the depth-first search ordering,
the new interleaving is guaranteed to be equivalent to one already checked,
and can be skipped.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.3\textwidth}p{0.25\textwidth}p{0.35\textwidth}}
	\begin{tabular}{l}
	$T_1$@$t_a$: \texttt{write~x} \\
	$T_2$@$t_b$: \texttt{read~y} \\
	$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_2$@$t_{c'}$: \texttt{read~x} \\
	$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_1$@$t_{a''}$: \texttt{write~x} \\
	$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\\
	\\
	(a) Original interleaving.
	&
	(b) Goal interleaving.
	&
	(c) Redundant interleaving.
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization.
	For clarity, threads are labeled here with numbers ($T_n$) and the transitions thereby with letters ($t_i$).}
	\label{fig:sleepsets}
\end{figure}
Landslide implements this check in {\tt equiv\_already\_explored()}.
This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which achieve optimality (i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented,
although to the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

@@

>>>>

@@

\subsubsection{Sleep set reduction}
\label{sec:landslide-sleepsets}

In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
For example, consider the execution sequence
in Figure~\ref{fig:sleepsets}(a).
After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
Figure~\ref{fig:sleepsets}(b).
However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
so from this branch,
it next tags $I_{c'1}$, causing
Figure~\ref{fig:sleepsets}(c).
to be tested next, which is regrettably equivalent to the original interleaving
($t_a$ and $t_b$ having been reordered all told, which are independent).
To recover our forgotten original purpose,
which was already fulfilled by testing the second interleaving,
DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
for an already-explored interleaving beginning with the target thread.
If such exists,
%then because of the depth-first search ordering,
the new interleaving is guaranteed to be equivalent to one already checked,
and can be skipped.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.3\textwidth}p{0.25\textwidth}p{0.35\textwidth}}
	\begin{tabular}{l}
	$T_1$@$t_a$: \texttt{write~x} \\
	$T_2$@$t_b$: \texttt{read~y} \\
	$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_2$@$t_{c'}$: \texttt{read~x} \\
	$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_1$@$t_{a''}$: \texttt{write~x} \\
	$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\\
	\\
	(a) Original interleaving.
	&
	(b) Goal interleaving.
	&
	(c) Redundant interleaving.
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization.
	For clarity, threads are labeled here with numbers ($T_n$) and the transitions thereby with letters ($t_i$).}
	\label{fig:sleepsets}
\end{figure}

Landslide implements this check in {\tt equiv\_already\_explored()}.
This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
Landslide's implementation differs from prior work by not explicitly tracking
sets of reordered threads and memory accesses they should expect to conflict with,
and instead identifies where the reduction should take place in subsequent iterations of DPOR.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which achieve optimality (i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented,
although to the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

### 2018-07-28 thesis/tm.tex
some mx verif results
more 10h expts
ok i think retry set impl is final, so write about it

\subsubsection{Adding possible abort codes}

@@

\subsubsection{Tracking abort sets}
\label{sec:tm-abortsets}

@@

The proof of soundness for this optimization is left to future work;
while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
its intersection with Landslide's other exploration features may cause unforeseen problems.
Accordingly, I implemented the optimization as conservatively as possible,
falling back to no pruning in cases where how to prune soundly is not clear
(for example, when a second pair of conflicting transactions involving a third thread
interleaves while a first abort set is already being tracked).

% TODO - rewrite the following example so that it only uses retry aborts
% ok so heres an example of why this is rly hard

% ==== thread 1 ====
% if (xbegin()) {
% 	xabort(0) // whatever, executed in a past branch to be aborting in the present
% } else {
% 	x++;
% }
%
% ==== thread 3 ====
% conflict_with_thread_2 = true;
%
% ==== thread 2 ====
% if (xbegin()) {
% 	conflict_with_thread_2 ++ ; // whatever, just conflict and fail w thr 3
% } else {
% 	x++; // conf with thr 1
% }
%
% so we execute this interleaving (T1 abort, T3, T2 abort)
% dpor sees the conflict between T2 and T1 and wants to tag T2 to run as transition(T1-abort)'s sibling to run
% it would use the "abort set" of (T2 [CONFLICT], T1 [EXPLICIT])
% but like, how do we know that thread 3 is necessary to reach this?
% and even, how do we know that it's valid to inject CONFLCIT in T2?
% you pretty much hafta reorder T3 (dpor doesn't see why you need to)
% but if you inject conflict in T2 without also making sure T3 executes "near" it, that might not necessarily be sound

@@

\subsection{Bugs}
\label{sec:tm-eval-bugs}

@@

\begin{table}[h]
	\begin{center}
		\footnotesize
		\begin{tabular}{cc|r|r}
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt htm2}
			& 2,1 & \cpu{16.30}	& \ints{22}	\\
			& 2,2 & \cpu{69.80}	& \ints{1970}	\\
			& 2,3 &\cpu{3510.00}	& \ints{104914}	\\
			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
			& 3,1 &\cpu{41.07}	& \ints{941}	\\
			& 3,2 &	\ETAdag{todo}	& \ETAdag{todo}	\\
			& 4,1 & \cpu{12699.65}	& \ints{344240}	\\
			\hline
			{\tt counter}
			& 2,1 & \cpu{5.40}	& \ints{10}	\\
			& 2,2 & \cpu{10.91}	& \ints{190}	\\
			& 2,3 & \cpu{134.21}	& \ints{3970}	\\
			& 2,4 & \cpu{3009.40}	& \ints{86950}	\\
			& 3,1 & \cpu{8.30}	& \ints{120}	\\
			& 3,2 & \cpu{2038.90}	& \ints{60606}	\\
			& 4,1 & \cpu{101.95}	& \ints{3006}	\\
			\hline
			{\tt swap}
			& 2,1 & \cpu{38.79}	& \ints{99}	\\
			& 2,2 & \cpu{10299.27}	& \ints{233396}	\\
			& 3,1 & \ETAdag{todo} & \ETAdag{todo}	\\
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{491.98}	& \ints{15056}	\\
			\hline
			{\tt avl\_fixed}
			& 2,1 & \cpu{572.49} 	& \ints{20420} 	\\
			& 2,2 & \ETAdag{todo}	& \ETAdag{todo} 	\\
			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	\\

@@

While verifying the soundness of Figure~\ref{fig:htm-fixed}'s atomicity protocol is a positive result,
the previous section's attempts at several larger real-world programs quickly suffered exponential explosion
for even small thread/iteration counts,
in one case failing to make any guarantee about even the minimum possible test configuration.
I explore two possible mitigation approaches.

@@

\begin{figure}[h]
	% TODO: syn hi
	\begin{center}
	\begin{tabular}{p{0.47\textwidth}p{0.5\textwidth}}
		\begin{tabular}{l}
			\texttt{\ctype{static int} num\_in\_section = 0;} \\
			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
			\texttt{~~~~num\_in\_section++;} \\
			\texttt{~~~~\flow{if} (!\call{\_xtest()})} \\
			\texttt{~~~~~~~~\call{thr\_yield}(\const{-1});} \\
			\texttt{~~~~\flow{assert}{}(num\_in\_section == \const{1}); } \\
			\texttt{~~~~num\_in\_section--;} \\
			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
			\texttt{\}} \\
		\end{tabular}
		&
		\begin{tabular}{ll}
			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
			\texttt{~~~~\flow{assert}{}(\call{\_xtest}());} \\
			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
			\texttt{\}} \\
		\end{tabular}
		\\
		\\
		(a) {\tt lock}(), tests mutual exclusion.
		&
		(b) {\tt lock\_fast}(), tests for no spurious aborts.
	\end{tabular}
	\end{center}
	\caption{Abstraction reduction test cases.}
	\label{fig:htm-lock}
\end{figure}

@@

		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + abort sets} \\

@@

			& 2,3 & \ETAdag{todo} & \ETAdag{todo} \\

@@

			& 2,4 & \ETAdag{todo} & \ETAdag{todo}	\\

@@

@@

		%% i mean ok xabort sets doesn't really help for ANY test tho
		%so the abort sets optimization does not apply to them.

@@

\subsubsection{Abort set reduction}

@@

(\sect{\ref{sec:tm-abortsets}}).

>>>>

\subsubsection{Adding possible abort codes}
\label{sec:txn-abort-codes}

@@

\subsection{Retry independence}
\label{sec:tm-retrysets}

Finally, I identified a specific pattern of transactional code where existing state-space reduction algorithms
will fail to identify and prune equivalent thread interleavings.
Figure~\ref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
In this program, each thread's transactional path conflicts with the other thread's abort path,
while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are pairwise independent.

\begin{figure}[h]
        \begin{center}
	% TODO syntax hilight
        \begin{tabular}{ll}
		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo, bar = \const{0}; \ctype{mutex\_t} m;}} \\
                \\
                \begin{tabular}{l}
                {\bf Thread 1} \\
                \hline
			\texttt{\flow{if} (\call{\_xbegin}() == \const{STARTED}) \{} \\
			\texttt{~~~~foo++;} \\
			\texttt{~~~~\call{\_xend}();} \\
			\texttt{\} \flow{else} \{} \\
			\texttt{~~~~\call{mutex\_lock}(\&m);} \\
			\texttt{~~~~\flow{assert}(foo + bar < \const{2});} \\
			\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
			\texttt{\}} \\
                \end{tabular}
                &
                \begin{tabular}{l}
                {\bf Thread 2} \\
                \hline
			\texttt{\flow{if} (\call{\_xbegin}() == \const{STARTED}) \{} \\
			\texttt{~~~~bar++;} \\
			\texttt{~~~~\call{\_xend}();} \\
			\texttt{\} \flow{else} \{} \\
			\texttt{~~~~\call{mutex\_lock}(\&m);} \\
			\texttt{~~~~\flow{assert}(foo + bar < \const{2});} \\
			\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
			\texttt{\}} \\
                \end{tabular}
        \end{tabular}
        \end{center}
	\caption{Motivating example for retry independence reduction.}
	\label{fig:retry-set-example}
\end{figure}

Ordinarily, in the state space subset which schedules thread 1 before thread 2,
there would be 4 combinations of each thread succeeding or aborting their respective transactions;
among those, at least one would show a memory conflict,
causing DPOR to explore into the other half of the state space which schedules thread 2 first,
where the same 4 combinations of transaction reuslts would be tested in the other order,
even though 2 are redundant under schedule reordering.
Consequently, na\"ively testing both success and retry aborts in both threads regardless of reordering
will unnecessarily execute % oops
both equivalent interleavings from each such pair;
to identify such equivalences,
DPOR must somehow remember which combination of {\tt xbegin} results led to the memory conflict in the first place.

While Figure~\ref{fig:retry-set-example}'s example may seem contrived
(what program's transactions would just give up and do no work if aborted?),
it is easy to imagine a larger transactional data structure,
two insertions into which might operate on disjoint nodes or array indices,
allowing simultaneous transactions to usually succeed,
while the more uncommon abort paths might take the opportunity to assert a full consistency check of all elements,
ultimately resulting in a similar conflict pattern.
Also, the evaluation will later show
that even programs with fully conflicting success/abort paths
may still exhibit some equivalent thread interleavings of this pattern
after their failure paths are split apart into smaller transitions by data-race preemption points
(\sect{\ref{sec:tm-eval-retrysets}}).

To solve this,
I extended Landslide
with an experimental feature called {\em retry set reduction},
so named after the analogous {\em sleep sets} from prior work
(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
if either or both sides of a memory conflict were part of a transaction,
it records a {\em retry set},
i.e., a pair of {\tt xbegin} results executed by the conflicting threads.
% ah-- not compatible after all!
%in addition to tracking the set of abort codes possible at each preemption point.
Later, when traversing that new branch,
Landslide remembers those threads' expected {\tt xbegin} results
and refuses to test any others
(unless DPOR separately found them, too, in conflict),
thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.}.

@@

The proof of soundness for this optimization is left to future work;
while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
its intersection with Landslide's other exploration features may cause unforeseen problems.
Accordingly, I implemented the optimization as conservatively as possible,
falling back to no pruning in cases where how to prune soundly is not clear
(for example, when a second pair of conflicting transactions involving a third thread
interleaves while a first retry set is already being tracked).
Note also that {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing this feature the wrong way at first.}%
),
so how to soundly prune under multiple abort codes
and/or when retry aborts are suppressed is left to future work.
For now, retry sets, true to their name,
are supported only under {\tt landslide -X},
and incompatible with both the {\tt -A} and {\tt -S} options.

@@

\subsection{Bug-finding}
\label{sec:tm-eval-bugs}

@@

\begin{table}[h]
	\begin{center}
		\footnotesize
		\begin{tabular}{cc|r|r}
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt htm2}
			& 2,1 & \cpu{16.30}	& \ints{22}	\\
			& 2,2 & \cpu{69.80}	& \ints{1970}	\\
			& 2,3 &\cpu{3510.00}	& \ints{104914}	\\
			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
			& 3,1 &\cpu{41.07}	& \ints{941}	\\
			& 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
			% [JOB 4] progress: 923822/5058149 brs (18.264030%), ETA 9d 8h 30m 38s (elapsed 9h 59m 55s)
			& 4,1 & \cpu{12699.65}	& \ints{344240}	\\
			\hline
			{\tt counter}
			& 2,1 & \cpu{5.40}	& \ints{10}	\\
			& 2,2 & \cpu{10.91}	& \ints{190}	\\
			& 2,3 & \cpu{134.21}	& \ints{3970}	\\
			& 2,4 & \cpu{3009.40}	& \ints{86950}	\\
			& 3,1 & \cpu{8.30}	& \ints{120}	\\
			& 3,2 & \cpu{2038.90}	& \ints{60606}	\\
			& 4,1 & \cpu{101.95}	& \ints{3006}	\\
			\hline
			{\tt swap}
			& 2,1 & \cpu{38.79}	& \ints{99}	\\
			& 2,2 & \cpu{10299.27}	& \ints{233396}	\\
			& 3,1 & \ETAdag{todo} & \ETAdag{todo}	\\
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{491.98}	& \ints{15056}	\\
			\hline
			{\tt avl\_fixed}
			& 2,1 & \cpu{572.49} 	& \ints{20420} 	\\
			& 2,2 & \ETAdag{todo}	& \ETAdag{todo} 	\\
			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	\\

@@

While verifying the soundness of Figure~\ref{fig:htm-fixed}'s atomicity protocol is a positive result,
the previous section's attempts at several larger real-world programs quickly suffered exponential explosion
for even small thread/iteration counts,
in one case failing to make any guarantee about even the minimum possible test configuration,
and in another failing to verify that a previously-found bug had been fixed.
I explore three possible mitigation approaches.

@@

\begin{figure}[h]
	\begin{center}
	\begin{tabular}{p{0.47\textwidth}p{0.5\textwidth}}
		\begin{tabular}{l}
			\texttt{\ctype{static int} num\_in\_section = 0;} \\
			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
			\texttt{~~~~num\_in\_section++;} \\
			\texttt{~~~~\flow{if} (!\call{\_xtest()})} \\
			\texttt{~~~~~~~~\call{thr\_yield}(\const{-1});} \\
			\texttt{~~~~\flow{assert}{}(num\_in\_section == \const{1}); } \\
			\texttt{~~~~num\_in\_section--;} \\
			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
			\texttt{\}} \\
		\end{tabular}
		&
		\begin{tabular}{ll}
			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
			\texttt{~~~~\flow{assert}{}(\call{\_xtest}());} \\
			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
			\texttt{\}} \\
		\end{tabular}
		\\
		\\
		(a) {\tt lock}(), tests mutual exclusion.
		&
		(b) {\tt lock\_fast}(), tests for no spurious aborts.
	\end{tabular}
	\end{center}
	\caption{Abstraction reduction test cases.}
	\label{fig:htm-lock}
\end{figure}

@@

		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + retry sets} \\

@@

			& 2,3 & \ETAdag{28h 6m} & \ETAdag{486966} \\ % certified causality ok, add 10h ok
			% [JOB 4] progress: 234307/486966 brs (48.115630%), ETA 18h 5m 37s (elapsed 9h 59m 50s)

@@

			& 2,4 & \cpu{13545.35} & \ints{384512}	\\ % certified causality ok

@@

			& 3,2 & \cpu{25434.85} & \ints{718690}	\\ % certified causality ok

@@

@@

\subsubsection{Retry set reduction}
\label{sec:tm-eval-retrysets}

@@

(\sect{\ref{sec:tm-retrysets}}).
